---
title: "interactive Immune MicroEnvironment - Univariate Report"
author: "iTIME Dev Team"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
header-includes:
  \usepackage{booktabs}
params:
  include_functions: NA # = input$printFunctions
  selected_marker: NA # = input$picked_marker
  picked_clinical: NA # = input$picked_clinical
  contingency_threshold: NA # = input$choose_cont_thresh
  total_cell_column: NA # = input$picked_total_cells
  modeling_reference: NA # = input$picked_modeling_reference
  boxplots: NA # = boxplot()
  contingency_table: NA # = contTable()
  frequency_table: NA # = freqTable()
  summary_table: NA # = summaryTable()
  chosen_model_stats: NA # = chosen_model_stats()
  cdf_plot: NA # = cdfplot()
  selected_univariate_model: NA # = input$selectedModel
  model_aic_table: NA # = aic_table()
  modelDescription: NA # = model_description()
---

<!--
install.packages(tinytex)
tinytex::install_tinytex()

\textcolor{display words}{color}
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(eval.after = "fig.cap")
library(kableExtra)
echo = params$include_functions
```

### Summary Table of Univariate Variables

Below in **Table 1** are the variables that were selected on iTIME.moffitt.org from the *summary* file and the *clinical* file in order to generate the tables and plots within.

```{r variable_table_chunk, echo=F}
variable_table = data.frame("Variable" = c("Include functions?",
                                           "Cell marker selected",
                                           "Clinical variable selected",
                                           "Threshold",
                                           "Column heading of total cell counts",
                                           "Clinical variable baseline"),
                            "Selection" = c(params$include_functions,
                                            params$selected_marker,
                                            params$picked_clinical,
                                            params$contingency_threshold,
                                            params$total_cell_column,
                                            params$modeling_reference),
                            "Held as on Server" = c("input$printFunctions",
                                                    "input$picked_marker",
                                                    "input$picked_clinical",
                                                    "input$choose_cont_thresh",
                                                    "input$picked_total_cells",
                                                    "input$picked_modeling_reference"), check.names = F)
kbl(variable_table,
    caption = paste("Variables selected in iTIME.Moffitt.org - Univariate Summary."),
    booktabs=T) %>%
  kable_styling(position="left", latex_options = c("hold_position"))
```

\clearpage
\newpage

### Contingency Table

```{r contingency_table_code, echo=echo, eval=F, size = "tiny"}
#Function to produce Contingency Table
contingency_table <- function(summary_clinical_merge, markers = markers,
                              clin_vars = clin_vars, percent_threshold = percent_threshold){
  #Maybe provide an error for multiple columns
  cells <- summary_clinical_merge %>% select(any_of(paste(markers)))
  assign("percent_threshold", percent_threshold, envir = .GlobalEnv)

  above <- function(x, percent_threshold){ifelse(x > percent_threshold,
                                                 paste0('Greater than ', percent_threshold, '%'),
                                                 paste0('Less than ', percent_threshold, '%'))}

  table <- cells %>% 
    mutate_all( ~ above(x = ., percent_threshold = as.numeric(percent_threshold))) %>%
    bind_cols(.,summary_clinical_merge %>% select(clin_vars)) %>%
    group_by(.[[paste(markers)]],.[[clin_vars]]) %>%
    summarize(n = n()) %>%
    pivot_wider(names_from = `.[[clin_vars]]`, values_from = n) %>%
    mutate_all(~replace_na(.,0))
  colnames(table)[1] = markers
  return(table)
}
```

```{r contingency_table_output, echo=F}
kbl(params$contingency_table,
    caption = paste("Contingency table of", params$selected_marker, "above and below the threshold of", params$contingency_threshold, " percent."),
    booktabs = TRUE) %>%
  kable_styling(position = "center", latex_options = c("hold_position"))
```

### Frequency Table

```{r frequency_table_code, echo=echo, eval = F, size = "tiny"}
#Function use to produce the frequency table
freq_table_by_marker <-
  function(summary_clinical_merge,
           clinical = clinical,
           markers = markers) {
    cells <-
      summary_clinical_merge %>% select(paste(clinical),any_of(paste( markers)))
    
    table <-
      cells %>% 
      mutate(`> 1%` = .[[paste( markers)]]>1,
             `> 2%` = .[[paste( markers)]]>2,
             `> 3%` = .[[paste( markers)]]>3,
             `> 4%` = .[[paste( markers)]]>4,
             `> 5%` = .[[paste( markers)]]>5) %>% 
      group_by(.[[paste(clinical)]]) %>%
      select(`> 1%`,`> 2%`,`> 3%`,`> 4%`,`> 5%`) %>%
      summarize_all( ~ sum(.))
    colnames(table)[1] = clinical
    
    return(table)
}
```

```{r frequency_table_output, echo=F}
kbl(params$frequency_table,
    caption = paste("Frequency table of", params$selected_marker, "by", params$picked_clinical, "."),
    booktabs = TRUE) %>%
  kable_styling(position = "center", latex_options = c("hold_position"))
```

### Summary Statistics Table

```{r summary_table_code, echo = echo, eval=F, size = "tiny"}
#Function used to produce the summary table
summary_table = function(summary_clinical_merged,
                         marker,
                         clinical,
                         merged){
  data = summary_clinical_merged %>% 
    select(paste(clinical), paste(merged), any_of(paste(marker)))
  colnames(data) = c("clinical","merged_var","marker")
  table = 
    data %>% 
    group_by(clinical) %>% 
    summarise(Min = min(marker),
              Median = median(marker),
              Mean = mean(marker),
              Max = max(marker),
              SD = sd(marker),
              Subjects = length(unique(merged_var)),
              Samples = length(marker))
  
  colnames(table)[1] = clinical
  return(table)
}

```

```{r summary_table_output, echo=F} 
kbl(params$summary_table,
    caption = paste("Summary table of", params$selected_marker, "by", params$picked_clinical, "showing the minimum, median, mean, max, and standard deviation as well as the number of unique subject and unique samples in the clinical and summary data frames."),
    booktabs = TRUE) %>%
  kable_styling(position = "center", latex_options = c("hold_position"))
```

\clearpage
\newpage

### Univariate Summary Plot

```{r univariate_plot_code, echo = echo, eval = F, size = "tiny"}
#Code within the server
message("Code within the server:")
cellvar <-  input$picked_marker
clinvar <- input$picked_clinical
colorscheme <- input$summaryPlotColors
data_table = summary_data_merged()

if(input$uni_transformation == "none"){
    thres = input$choose_cont_thresh
} else if(input$uni_transformation == "sqrt_transform"){
    data_table[,cellvar] = sqrt(data_table[,cellvar])
    thres = sqrt(as.numeric(input$choose_cont_thresh))
} else if(input$uni_transformation == "log2_transform"){
    data_table[,cellvar] = log2(data_table[,cellvar]+0.0001)
    thres = log2(as.numeric(input$choose_cont_thresh)+0.0001)
} else if(input$uni_transformation == "logit_transform"){
    p = (data_table[,cellvar]/100)+0.0001
    data_table[,cellvar] = log10(p/(1-p))
    tmp = (as.numeric(input$choose_cont_thresh)/100) + 0.0001
    thres = log10(tmp/(1-tmp))
}
plots = summary_plots_fn(data_table, clinvar, cellvar, colorscheme, thres)
plots[[as.integer(input$summaryPlotType)]]

#Function to produce all of the summary plots
summary_plots_fn <- function(datatable, clinvar, cellvar, colorscheme, threshold){
  box_p <- ggplot(datatable, aes(x=get(clinvar), y=get(cellvar), fill=get(clinvar))) + 
    geom_boxplot() +
    xlab(str_to_title(clinvar)) + ylab(gsub("_", " ", str_to_title(cellvar))) +
    labs(fill=str_to_title(clinvar)) + theme_classic(base_size = 20) +
    viridis::scale_fill_viridis(option = colorscheme, discrete = TRUE) + 
    geom_hline(yintercept = as.numeric(threshold), size = 1.25, 
               linetype = "twodash", color = 'red') + 
    theme(legend.position = 'none')
  
  violin_p <- ggplot(datatable, aes(x=get(clinvar), y=get(cellvar), fill=get(clinvar))) + 
    geom_violin() +
    xlab(str_to_title(clinvar)) + ylab(gsub("_", " ", str_to_title(cellvar))) +
    labs(fill=str_to_title(clinvar)) + theme_classic(base_size = 20) +
    viridis::scale_fill_viridis(option = colorscheme, discrete = TRUE) + 
    geom_hline(yintercept = as.numeric(threshold), size = 1.25, 
               linetype = "twodash", color = 'red') + 
    theme(legend.position = 'none')
  
  hist_p <- ggplot(datatable, aes(x=get(cellvar), color=get(clinvar))) + 
    geom_histogram(position='stack', fill = 'white') + facet_wrap(get(clinvar)~., nrow = 1) +   
    xlab(str_to_title(gsub("_", " ", cellvar))) + ylab("Count") +
    labs(fill=str_to_title(clinvar)) + theme_classic(base_size = 20) +
    viridis::scale_color_viridis(option = colorscheme, discrete = TRUE) + 
    theme(legend.position = 'none') + 
    geom_vline(xintercept = as.numeric(threshold), 
               size = 1.25, linetype = "twodash", color = 'red')
  
  if(is.character(datatable[[clinvar]])){
    scatter_p <- ggplot(datatable, aes(x=get(clinvar), y=get(cellvar), color=get(clinvar))) +
      geom_point() +
      xlab(str_to_title(clinvar)) + ylab(gsub("_", " ", str_to_title(cellvar))) +
      labs(color=str_to_title(clinvar)) + theme_classic(base_size = 20) +
      viridis::scale_color_viridis(option = colorscheme, discrete=TRUE)
  }
  else{
    scatter_p <- ggplot(datatable, aes(x=get(clinvar), y=get(cellvar), color=get(clinvar))) +
      geom_point() +
      xlab(str_to_title(clinvar)) + ylab(gsub("_", " ", str_to_title(cellvar))) +
      labs(color=str_to_title(clinvar)) + theme_classic(base_size = 20) +
      viridis::scale_color_viridis(option = colorscheme, discrete=FALSE)    
  }
  
  summ_plots <- list(box_p, violin_p, hist_p, scatter_p)
  
  return(summ_plots)
}
```

```{r univariate_plot_output, echo=F, fig.height=4, fig.width=8, fig.cap = paste("Summary plot of ", params$selected_marker, " separated by ", params$picked_clinical,". The red line is at the threshold level in Table 1.", sep="")}
params$boxplots
```

\clearpage
\newpage

### Cumulative Distribution Function Plot

```{r cdf_plot_code, echo = echo, eval = F, size = "tiny"}
#Code within the server
marker = input$picked_marker
data_table = summary_data_merged()
CDF_plots(summary_data_merge = data_table, markers = substr(marker, 9, nchar(marker)))

#Function used to produce the CDF plot
CDF_plots = function(summary_data_merged = summary_data_merged, markers = markers){
  sample_stats = summary_data_merged %>% select(grep('Total', colnames(.)), markers) %>%
    pivot_longer(cols = 2:ncol(.), values_to = 'Count', names_to = 'Marker') %>%
    group_by(Marker) %>%
    summarize(prob0 = mean(Count == 0, na.rm = TRUE),
              Avg_p = mean(Count/`Total Cells`, na.rm = TRUE),
              Avg_Count = mean(Count, na.rm = TRUE),
              Avg_Total = round(mean(`Total Cells`, na.rm = TRUE)))
   
  cdfs = summary_data_merged %>%  select(grep('Total', colnames(.)), markers) %>% 
    pivot_longer(cols = 2:ncol(.), names_to = 'Marker', values_to = 'Count') %>%
    mutate(ecdf = ecdf(Count)(Count)) %>%
    mutate(Poisson =  ppois(q = Count, 
                            lambda = sample_stats$Avg_Count),
           Binomial = pbinom(q = Count, 
                             size = round(sample_stats$Avg_Total),
                             prob = sample_stats$Avg_p,
           ),
           `ZI Poisson` =  pzipois(q = Count, 
                                   lambda = sample_stats$Avg_Count,
                                   pstr0 = sample_stats$prob0
           ),
           `ZI Binomial` = pzibinom(q = Count, 
                                    size = round(sample_stats$Avg_Total),
                                    prob = sample_stats$Avg_p,
                                    pstr0 = sample_stats$prob0
           ),
           `Negative Binomial` =  pnbinom(q = Count, 
                                          size = round(sample_stats$Avg_Count),
                                          prob = 1 - sample_stats$prob0
           ),
           `Beta Binomial` =  pbetabinom(q = Count, 
                                         size = round(sample_stats$Avg_Total),
                                         prob = sample_stats$Avg_p,
                                         rho = sample_stats$prob0)
           
    ) %>% 
    pivot_longer(col = 5:ncol(.), values_to = 'CDF', names_to = 'Distribution')

  cdfs = cdfs %>% 
    mutate(family = ifelse(Distribution %in% c('Poisson', 'ZI Poisson', 'Negative Binomial'),
                           'Poisson', 'Binomial'),
           Distribution = factor(Distribution))

  binomial_plot = cdfs %>% 
    ggplot(aes(x = Count, y = ecdf, color = 'Empirical')) + 
    geom_line(aes(color = 'Empirical'), color = 'black') + 
    geom_line(aes(x = Count, y = CDF, color = Distribution, linetype = family)) + theme_bw() + 
    theme(axis.text.x = element_blank(),
          axis.ticks = element_blank(),
          axis.title.x = element_blank(),
          legend.position = "right", #c(0.8, 0.3),
          axis.title.y = element_text(size = 16),
          axis.text.y = element_text(size = 16),
          strip.text = element_text(size=16),
          legend.text = element_text(size = 16),
          legend.title = element_text(size = 16)) + 
    labs(color = 'Distribution') + scale_linetype_manual(values = c("solid", "longdash"))

  return(binomial_plot)
}
```

```{r cdf_plot_output, echo=F, fig.height=4, fig.width=8, fig.cap = paste("Cumulative distribution functions for both the binomial and Poisson families for ", params$selected_marker, ". The default model selected in iTIME.Moffitt.org is the negative binomial model.", sep="")}
params$cdf_plot
```

\pagebreak

### Betabinomial Model Statistics

```{r model_statistics_code, echo = echo, eval=F, size = "tiny"}
#Function to determine if there are repeated measures for individuals
model_checked_repeated = function(summary_data_merged,markers,
                                  Total, clin_vars, reference,
                                  choose_clinical_merge, digits=4){
  round4 = function(x){return(round(x,digits=digits))}
  markers = markers[markers %in% colnames(summary_data_merged)]
  #Check whether or not any subject has multiple samples
  if(any(table(summary_data_merged[[choose_clinical_merge]])>1)){
    return(
      models_repeated_measures(summary_data_merged, markers,
                               Total, clin_vars, reference, 
                               choose_clinical_merge, digits))
  }else{
    return(
      models(summary_data_merged, markers,Total, clin_vars, reference, digits))
  }
}

#function to model without repeated measures
models = function(summary_data_merged, markers,
                  Total, clin_vars, reference,digits=4){
  out = list()
  levels = unique(summary_data_merged[[clin_vars]])
  tmp = summary_data_merged %>% select(Total, markers, clin_vars) %>%
    mutate(clin_vars = factor(.[[clin_vars]], 
                                   levels = c(reference,
                                              levels[levels != reference]
                                              )))
  round4 = function(x){return(round(x,digits=digits))}

  #Fit Betabinomial model
  model_fit_bb = try(VGAM::vglm(cbind(tmp[[markers]], tmp[[Total]] - tmp[[markers]]) ~ tmp$clin_vars, 
                             betabinomial(zero = 2), data = tmp), silent = TRUE)
  
  out$models = list('Beta Binomial' = model_fit_bb
  return(out)
}

#function to model with repeated measures of individuals
models_repeated_measures = function(summary_data_merged, markers,
                  Total, clin_vars, reference, choose_clinical_merge, digits = 4){
  out = list()
  round4 = function(x){return(round(x,digits=digits))}
  levels = unique(summary_data_merged[[clin_vars]])
  tmp = summary_data_merged %>% select(choose_clinical_merge, 
                                       Total, markers, clin_vars) %>%
    mutate(clin_vars = factor(.[[clin_vars]], 
                              levels = c(reference,
                                         levels[levels != reference]
                              )),
           id = factor(.[[choose_clinical_merge]]))
  
  #Fit Betabinomial model
  model_fit_bb = try(GLMMadaptive::mixed_model(
    fixed = cbind(tmp[[markers]], tmp[[Total]] - tmp[[markers]]) ~ 
      tmp$clin_vars, random = ~ 1|id, 
    family = beta.binomial(), data = tmp), silent = TRUE)
  
  out$models = list('Beta Binomial' = model_fit_bb)
  return(out)
}

#server code to run model
model_list = reactive({
    validate(need(input$picked_clinical !="", "Please select a clinical variable....."),
             need(ncol(summary_data_merged()) > 0, "Please upload clinical and summary data....."),
             need(input$picked_marker !="", "Please pick a marker....."),
             need(input$picked_total_cells !="", "Please select column with total cell count....."),
             need(input$picked_modeling_reference !="", "Please wait while statistics are computed....."))
    marker = input$picked_marker
    marker = substr(marker, 9, nchar(marker))
    marker = c(marker, gsub("\\ Positive\\ ", "\\ ", marker))
    suppressWarnings({
        df = model_checked_repeated(summary_data_merged = summary_data_merged(), markers = marker,
                                    Total = input$picked_total_cells, clin_vars = input$picked_clinical, reference = input$picked_modeling_reference,
                                    choose_clinical_merge = input$clinical_merge) #assuming IDs are merging variable (patientID, subjectID, etc)
    })
    return(df)
})

#server code to produce summary stats table
chosen_model_stats = reactive({
    validate(need(model_list(), "Please wait while things finish loading....."))
    models1 = model_list()
    df = models1$models[["Beta Binomial"]]
    if(class(df)=="character"){
        df1 = data.frame(df)
    } else if(class(df)=="MixMod"){
        df1 = summary(df)$coef_table
        df1 = data.frame(Terms = gsub("tmp\\$clin_vars", "", row.names(df1)),
                         df1, check.names = F)
    }else{
        df = df %>% summary() %>% coefficients()#input$selectedModel
        df1 = data.frame(Terms = gsub("tmp\\$clin_vars", "", row.names(df)),
                         df, check.names = F)
        df1 = df1[-2,]
    }
    
    return(df1)
})
```

```{r model_statistics_output, echo=F}
kbl(params$chosen_model_stats,
    caption = paste("Summary table of betabinomial model fit for", params$selected_marker, "by", params$picked_clinical, "using", params$modeling_reference,
                    "as the reference level of", params$picked_clinical, "."),
    booktabs = TRUE) %>%
  kable_styling(position = "center", latex_options = c("hold_position"))
```

`r params$modelDescription`






















