---
title: "summary_plots_iTime"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load `r tidyverse` for data frame manipulation.
```{r, message=F}
library('tidyverse')
```

Load data from .Rdata file.
```{r}
load("../data/example_data.RData")
```

Clean column names for the image summary data and the clinical data.
```{r}
summary_clean <- janitor::clean_names(summary)
clinical_clean <- janitor::clean_names(clinical)
```

Join the image summary data and the clinical data by the filename of the image ("Image tag").
```{r}
summary_clinical <- left_join(clinical_clean, summary_clean, by="image_tag")
```

Show columns available for selection.
```{r}
cat("CLINICAL VARIABLES:\n")
names(clinical_clean)
cat("\n")
cat("CELL TYPES:\n")
names(summary_clean)
```

Specify user-selected clinical column.
Specify user-selected cell type column.
```{r}
clinical_var <- 'race'
celltype_var <- 'cd3_opal_570_positive_cells'
```

Generate boxplot.
```{r}
box_p <- ggplot(summary_clinical, aes(x=get(clinical_var), y=get(celltype_var), fill=get(clinical_var))) + 
  geom_boxplot() +
  xlab(str_to_title(clinical_var)) + ylab(gsub("_", " ", str_to_title(celltype_var))) +
  labs(fill=str_to_title(clinical_var))

box_p
```
